### 2022-07-06 ehoh/ehoh.tex
intro
more
martin comments 1

\looseness=-1
A central aspect of our work is a set of techniques we call
\emph{prefix optimization}. Taking a traditional look at higher-order terms, they contain twice as many proper
subterms as first-order terms; for example,
$\cst{f}\;(\cst{g}\;\cst{a})\;\cst{b}$ contains not only the ``argument'' subterms
$\cst{g}\;\cst{a}$, $\cst{a}$, $\cst{b}$ but also the ``prefix'' subterms
$\cst{f}$, $\cst{f}\;(\cst{g}\;\cst{a})$, $\cst{g}$.
\begin{rep}Many operations, including superposition and rewriting, require
traversing all subterms of a term.\end{rep}
Using the optimization, the prover traverses subterms recursively in a
first-order fashion, considering all the prefixes of a given subterm
together. %, at little additional cost.
%
Our experiments (Sect.~\ref{sec:ehoh:evaluation}) show that Ehoh is
almost as fast as E on first-order problems and can also prove
higher-order problems that do not require synthesizing
$\lambda$-terms. In Chapter~\ref{ch:ehoh2} we show how to add support for
$\lambda$-terms and higher-order unification.

@@

\looseness=-1
Our logic is a variant of the intensional $\lambda$-free Boolean-free
higher-order logic (\lfhol{}) described by Bentkamp et al.\
%, Blanchette, Cruanes, and Waldmann
\cite[Sect.~2]{bbcw-21-lfho}, which could also be called ``applicative
first-order logic.'' In the spirit of FOOL \cite{kotelnikov-16-fool}, we
%% kotelnikov-15-fool
extend the syntax of this logic by making formulas a special case of terms (without adding logical symbols to the signature),
and its semantics by interpreting the Boolean type $o$ as a domain of
cardinality~2. Functional extensionality, the property that two functions are equal if they
always return the same value when given the same arguments, can be obtained by
adding suitable axioms can be obtained by adding suitable axioms
\cite[Sect.~3.1]{bbcw-21-lfho}.

@@

\looseness=-1
In \lfhol{}, a type signature is simply a type~$\tau$, in which the
type constructor $\to$ can be nested---e.g., $(\iota \to \iota)\allowbreak
\to \iota.$
%
A natural way to represent such types is to mimic their recursive
structure using a tagged union. However, this leads to memory fragmentation;
a simple operation such as querying the type of a function's $i$th
argument would require dereferencing $i$ pointers. We prefer a
flattened representation, in
which a type $\tau_1 \rightarrow \cdots \rightarrow \tau_n \rightarrow \iota$
is represented by a single node labeled with ${\rightarrow}$ and pointing to
the array $(\tau_1,\dots,\tau_n,\iota).$

@@

In first-order logic, the arity of a variable is always 0, and the arity of a
symbol is given by its type signature.
In higher-order logic, variables may have function type and be applied, and
symbols can be applied to fewer arguments than specified by their type
signatures. A natural representation of \lfhol{} terms as tagged unions
would distinguish between variables~$x$, symbols~$\cst{f}$, and binary
applications $s \; t.$ However, this scheme suffers from memory
fragmentation and linear-time access, as with the representation of types,
affecting performance on purely or mostly first-order problems. Instead, we
propose a flattened representation, as a generalization of E's existing data
structures: Allow arguments to variables, for symbols let \verb|arity| be
the number of \begin{conf}\emph{actual} arguments\end{conf}\begin{rep}actual
arguments, %as opposed to the declared arity,
and rename the field \verb|num_args|\end{rep}. This representation, often called
``spine notation,'' % \cite{cervesato-pfenning-2003},
is isomorphic to the standard definition of higher-order terms with binary
application. It is employed in various higher-order reasoning systems,
including Leo-III \cite{sb-21-leo3} and
Zipperposition \cite{sc-15-simon-phd,sc-supind-17}.

@@

\subsection{Unification}
We present our unification procedure as a \begin{rep}nondeterministic \end{rep}transition system
that generalizes Baader and Nipkow \cite{bn-98-tr-and-all-that}.
A unification problem consists of a finite set~$\mathit{S}$ of unification constraints $s_i
\unif t_i$, where $s_i$ and $t_i$ are of the same type.
A problem is in \emph{solved form} if it has the form $\{ x_1 \unif t_1,\allowbreak
\ldots, x_n \unif t_n \}$, where the $x_i$'s are distinct and do not occur in
the $t_j$'s. The corresponding unifier is
$\{ x_1 \mapsto t_1,\allowbreak \ldots, x_n \mapsto t_n \}.$
%
The transition rules attempt to bring the input constraints into solved form.
\begin{rep}They can be applied in any order and eventually reach a normal form, which is
either an idempotent MGU expressed in solved form or the special
value $\bot$, denoting unsatisfiability of the constraints.\end{rep}

@@

    The proof of the problem's unsolvability if rule \textsf{Clash} or
    \textsf{OccursCheck} is applicable carries over from Baader and Nipkow.
    For \textsf{ClashTypeX}, the justification is that $\sigma(x \;
    \tuple{s}{m}) = \sigma(u \; \tuple{t}{m})$ is possible only if
    $\sigma(x) = \sigma(u)$, which requires $x$ and $u$ to have the same
    type. Similarly, for \textsf{ClashLenXF}, if
    $\sigma(x \; \tuple{s}{m}) = \sigma(\cst{f} \; \tuple{t}{n})$ with
    $m > n$, we must have $\sigma(x \; \tuple{s}{{m-n}}) =
    \sigma(x) \; \sigma(s_1) \, \dots \, \sigma(s_{m-n}) = \cst{f}$,
    which is impossible.
  \end{proof}
\newpage
  \begin{lemma}\label{lemma:ehoh:unif-normal-imp-solved}
    If $\mathit{S}$ is a normal form, then $\mathit{S}$ is in solved form.
  \end{lemma}
  \begin{proof}
    Consider an arbitrary unification constraint $s \unif t \in \mathit{S}.$ We show
    that in all but one cases, a rule is applicable, contradicting the
    hypothesis that $\mathit{S}$ is a normal form. In the remaining case, $s$ is a
    solved variable in $\mathit{S}$.

@@

The superposition rules were introduced in Sect.~\ref{sec:pre:rules}. For completeness,
we repeat them with slightly simplified notation, as we do not repeat side conditions:

@@

For example, let
$\cst{f} \; (\cst{g} \; \cst{a} \; \cst{b}) \eq \cst{f} \; (\cst{h} \; \cst{g} \; \cst{b})$
be the given clause,
and suppose that $x \; \cst{a} \eq \cst{h} \; x$ is indexed.
Under context $\cst{f}\>[\phantom{.}]$, Ehoh considers the subterms
$\cst{g} \; \cst{a} \; \cst{b}$ and $\cst{h} \; x \; \cst{b}$. It finds the
prefix $\cst{g} \; \cst{a}$ of $\cst{g} \; \cst{a} \; \cst{b}$ in the
%
tree, with $\sigma = \{x \mapsto\nobreak \cst{g}\}$.
The prefix $\cst{h} \; \cst{g}$ of $\cst{h} \; \cst{g} \; \cst{b}$ matches the
indexed equation's right-hand side $\cst{h} \; x$ using the same substitution,
and the remaining argument in both subterms,~$\cst{b}$, is identical.
\begin{rep}Ehoh concludes that the given clause is redundant.\end{rep}

@@

\noindent
where $\textsf{sk}$ is a fresh Skolem symbol, and
$J$ is the largest subset of $\{1, \ldots, n$\} such that $x_j = y_j$ for every $j \in J$.
We denote the subsequence of $\tuple{x}{n}$ indexed by $J$ by $\tuple{x}{J}$.
Moreover, we require that
$x_i \neq y_i$, all variables in $\tuple{x}{K} \cdot \tuple{y}{K}$ are distinct,
where $K = \{1, \ldots, n\} \setminus J$,
and neither $\tuple{x}{K}$ nor $\tuple{y}{K}$ shares variables with $\tuple{x}{J}$.
For example, given $\cst{add} \; a \; b \noteq \cst{add} \; a \; b' \mathrel\lor a \eq b'$,
\infname{IR} can derive the existence of the inverse
$\cst{sk}_1$ characterized by $\cst{sk}_{1\!} \; (\cst{add} \; a \; b) \; a \eq b$.
%As a second example, from the clause
%$\cst{cons} \; x \; \mathit{xs} \noteq \cst{cons} \; y \; \mathit{ys}
%\mathrel\lor x \eq y$, we can derive
%$\cst{sk}_2 \; (\cst{cons} \; x \; \mathit{xs}) \eq x$.

>>>>

\looseness=-1
A central aspect of our work is a set of techniques we call
\emph{prefix optimization}. Taking a traditional look at higher-order terms, they contain twice as many proper
subterms as first-order terms; for example,
$\cst{f}\;(\cst{g}\;\cst{a})\;\cst{b}$ contains not only the ``argument'' subterms
$\cst{g}\;\cst{a}$, $\cst{a}$, $\cst{b}$ but also the ``prefix'' subterms
$\cst{f}$, $\cst{f}\;(\cst{g}\;\cst{a})$, $\cst{g}$.
\begin{rep}Many operations, including superposition and rewriting, require
traversing all subterms of a term.\end{rep}
The prefix optimization allows the prover to traverse subterms recursively in a
first-order fashion, and simultaneously consider the prefixes of a given subterm
at almost no additional cost.
%
Our experiments (Sect.~\ref{sec:ehoh:evaluation}) show that Ehoh is
almost as fast as E on first-order problems and can also prove
higher-order problems that do not require synthesizing
$\lambda$-terms. In Chapter~\ref{ch:ehoh2} we show how to add support for
$\lambda$-terms and higher-order unification.

@@

\looseness=-1
Our logic is a variant of the intensional $\lambda$-free Boolean-free
higher-order logic (\lfhol{}) described by Bentkamp et al.\
%, Blanchette, Cruanes, and Waldmann
\cite[Sect.~2]{bbcw-21-lfho}, which could also be called ``applicative
first-order logic.'' In the spirit of FOOL \cite{kotelnikov-16-fool}, we
%% kotelnikov-15-fool
extend the syntax of this logic by making formulas a special case of terms (without adding logical symbols to the signature),
and its semantics by interpreting the Boolean type $o$ as a domain of
cardinality~2. Functional extensionality, the property that two functions are equal if they
always return the same value when given the same arguments, can be obtained by
adding suitable axioms
\cite[Sect.~3.1]{bbcw-21-lfho}.

@@

\looseness=-1
In \lfhol{}, function types are built using the type constructor $\to$, which can be
arbitrarily nested---e.g., $(\iota \to \iota)\allowbreak
\to \iota.$
%
A natural way to represent such types is to mimic their recursive
structure using a tagged union. However, this leads to memory fragmentation;
a simple operation such as querying the type of a function's $i$th
argument would require dereferencing $i$ pointers. We prefer a
flattened representation, in
which a type $\tau_1 \rightarrow \cdots \rightarrow \tau_n \rightarrow \iota$
is represented by a single node labeled with ${\rightarrow}$ and pointing to
the array $(\tau_1,\dots,\tau_n,\iota).$

@@

In first-order logic, the arity of a variable is always 0, and the arity of a
symbol is given by its type signature.
In higher-order logic, variables may have function type and be applied, and
symbols can be applied to fewer arguments than specified by their type
signatures. A natural representation of \lfhol{} terms as tagged unions
would distinguish between variables~$x$, symbols~$\cst{f}$, and binary
applications $s \; t.$ However, this scheme suffers from memory
fragmentation and linear-time access, as with the representation of types,
affecting performance on purely or mostly first-order problems. Instead, we
propose a flattened representation, as a generalization of E's existing data
structures: Allow arguments to variables, for symbols let \verb|arity| be
the number of \begin{conf}\emph{actual} arguments\end{conf}\begin{rep}actual
arguments, %as opposed to the declared arity,
and rename the field to \verb|num_args|\end{rep}. This representation, often called
``spine notation,'' % \cite{cervesato-pfenning-2003},
is isomorphic to the standard definition of higher-order terms with binary
application. It is employed in various higher-order reasoning systems,
including Leo-III \cite{sb-21-leo3} and
Zipperposition \cite{sc-15-simon-phd,sc-supind-17}.

@@

\subsection{Unification}
We present our unification procedure as a \begin{rep}nondeterministic \end{rep}transition system
that generalizes the one of Baader and Nipkow \cite{bn-98-tr-and-all-that}.
A unification problem consists of a finite set~$\mathit{S}$ of unification constraints $s_i
\unif t_i$, where $s_i$ and $t_i$ are of the same type.
A problem is in \emph{solved form} if it has the form $\{ x_1 \unif t_1,\allowbreak
\ldots, x_n \unif t_n \}$, where the $x_i$'s are distinct and do not occur in
the $t_j$'s. The corresponding unifier is
$\{ x_1 \mapsto t_1,\allowbreak \ldots, x_n \mapsto t_n \}.$
%
The transition rules attempt to bring the input constraints into solved form.
\begin{rep}They can be applied in any order and eventually reach a normal form, which is
either an idempotent MGU expressed in solved form or the special
value $\bot$, denoting unsatisfiability of the constraints.\end{rep}

@@

    The proof of the problem's unsolvability if rule \textsf{Clash} or
    \textsf{OccursCheck} is applicable carries over from Baader and Nipkow.
    For \textsf{ClashTypeX}, the justification is that $\sigma(x \;
    \tuple{s}{m}) = \sigma(u \; \tuple{t}{m})$ is possible only if
    $\sigma(x) = \sigma(u)$, which requires $x$ and $u$ to have the same
    type. Similarly, for \textsf{ClashLenXF}, if
    $\sigma(x \; \tuple{s}{m}) = \sigma(\cst{f} \; \tuple{t}{n})$ with
    $m > n$, we must have $\sigma(x \; \tuple{s}{{m-n}}) =
    \sigma(x) \; \sigma(s_1) \, \dots \, \sigma(s_{m-n}) = \cst{f}$,
    which is impossible.
  \end{proof}

  \begin{lemma}\label{lemma:ehoh:unif-normal-imp-solved}
    If $\mathit{S}$ is a normal form, then $\mathit{S}$ is in solved form.
  \end{lemma}
  \begin{proof}
    Consider an arbitrary unification constraint $s \unif t \in \mathit{S}.$ We show
    that in all but one cases, a rule is applicable, contradicting the
    hypothesis that $\mathit{S}$ is a normal form. In the remaining case, $s$ is a
    solved variable in $\mathit{S}$.

@@

\begin{sloppypar}
  The rules of the superposition calculus were introduced in Sect.~\ref{sec:pre:rules}. For completeness,
  we repeat them with slightly simplified notation, as we do not repeat side conditions:
\end{sloppypar}

@@

For example, let
$\cst{f} \; (\cst{g} \; \cst{a} \; \cst{b}) \eq \cst{f} \; (\cst{h} \; \cst{g} \; \cst{b})$
be the given clause,
and suppose that $x \; \cst{a} \eq \cst{h} \; x$ is indexed.
Under context $\cst{f}\>[\phantom{.}]$, Ehoh considers the subterms
$\cst{g} \; \cst{a} \; \cst{b}$ and $\cst{h} \; \cst{g} \; \cst{b}$. It finds the
prefix $\cst{g} \; \cst{a}$ of $\cst{g} \; \cst{a} \; \cst{b}$ in the
%
tree, with $\sigma = \{x \mapsto\nobreak \cst{g}\}$.
The prefix $\cst{h} \; \cst{g}$ of $\cst{h} \; \cst{g} \; \cst{b}$ matches the
indexed equation's right-hand side $\cst{h} \; x$ using the same substitution,
and the remaining argument in both subterms,~$\cst{b}$, is identical.
\begin{rep}Ehoh concludes that the given clause is redundant.\end{rep}

@@

\noindent
where $\textsf{sk}$ is a fresh Skolem symbol, and
$J$ is the largest subset of $\{1, \ldots, n$\} such that $x_j = y_j$ for every $j \in J$.
We denote the subsequence of $\tuple{x}{n}$ indexed by $J$ by $\tuple{x}{J}$.
Moreover, we require that
$x_i \neq y_i$, all variables in $\tuple{x}{K} \cdot \tuple{y}{K}$ are distinct,
where $K = \{1, \ldots, n\} \setminus J$,
and neither $\tuple{x}{K}$ nor $\tuple{y}{K}$ shares variables with $\tuple{x}{J}$.
For example, given $\cst{add} \; a \; b \noteq \cst{add} \; a \; b' \mathrel\lor b \eq b'$,
\infname{IR} can derive the existence of the inverse
$\cst{sk}_1$ characterized by $\cst{sk}_{1\!} \; (\cst{add} \; a \; b) \; a \eq b$.
%As a second example, from the clause
%$\cst{cons} \; x \; \mathit{xs} \noteq \cst{cons} \; y \; \mathit{ys}
%\mathrel\lor x \eq y$, we can derive
%$\cst{sk}_2 \; (\cst{cons} \; x \; \mathit{xs}) \eq x$.

### 2022-07-06 pre/pre.tex
martin comments 1

Even though we do not formally define the semantics of logic, we assume the natural
extensions of domain, valuation, interpretation, and model (as defined by Fitting
\cite{mf-1996-fol}) from unsorted to many-sorted logic. The models we consider are
\emph{normal}, i.e., they interpret $\eq$ as an equality relation. Usual notions
of (un)satisfiability and (in)validity are assumed. We write
$\ourmodel~\models_\xi~N$ to denote that a model $\ourmodel$ satisfies a clause
set $N$, for a variable assignment $\xi$. If $\ourmodel$ is a model of $N$
(i.e., $\ourmodel$ satisfies it under every variable assignment), we simply write
$\ourmodel \models N.$  Overloading notation, we write $M \models N$ to denote that
$M$ \emph{entails} $N$, i.e., that every model of $M$ is a model of $N.$

\looseness=-1
A {\em position} in a term is a tuple of natural numbers, with $\varepsilon$
denoting the empty tuple. {\em Subterms} and positions are inductively defined
as follows. The term $t$ is a subterm of itself at position $\varepsilon$. If
$s$ is a subterm of $t_i$ at position $p$, then $s$ is a subterm of
$\cst{f}(\tuplen{t})$ at position $i.p$. A {\em context} is a term with zero or
more subterms replaced by a hole $\square$. We write $C[\overline{u}_n]$ for the
term resulting from filling in the holes of a context $C$ with the terms
$\overline{u}_n$, from left to right. We say a term is {\em ground} if it has no
variables. By $u[s]$ we denote that $s$ is a subterm of $u$.

\looseness=-1
{\em Substitutions} $\sigma, \varrho, \ldots$ are total mappings from variables to
terms of the same type. Substitutions map only finitely many variables to terms
other than the variable itself. This is denoted as $\{ x_1 \mapsto t_1, \ldots,
x_n \mapsto t_n \}$ where $x_i$ are variables that are not mapped to themselves.
Applying a substitution $\sigma$ to a term $t$, denoted $\sigma(t)$, results in
replacing all mapped variables by the corresponding terms.
The composition $\varrho\sigma$ of substitutions is defined by
$\left(\varrho\sigma\right)t=\varrho\left(\sigma (t)\right)$. We call a substitution $\sigma$ {\em grounding}
if $\sigma(x) \not= x$ implies that $\sigma(x)$ is ground.

@@

\[
\namedsimp{RN}{s \eq t \qquad u \noteq v \llor D}
              {s \eq t \qquad \sigmacl{u[p \leftarrow \sigma(t)] \eq v \llor C \llor D}}
\]
where $\sigma = \mgu(u|_p, s)$ and $\sigma(s) \succ \sigma(t)$.

>>>>

Even though we do not formally define the semantics of logic, we assume the natural
extensions of domain, valuation, interpretation, and model (as defined by Fitting
\cite{mf-1996-fol}) from unsorted to many-sorted logic. The models we consider are
\emph{normal}, i.e., they interpret $\eq$ as an equality relation. Usual notions
of (un)satisfiability and (in)validity are assumed. We write
$\ourmodel~\models_\xi~N$ to denote that a model $\ourmodel$ satisfies a formula
set $N$, for a variable assignment $\xi$. If $\ourmodel$ is a model of $N$
(i.e., $\ourmodel$ satisfies it under every variable assignment), we simply write
$\ourmodel \models N.$  Overloading notation, we write $M \models N$ to denote that
$M$ \emph{entails} $N$, i.e., that every model of $M$ is a model of $N.$

\looseness=-1
A {\em position} in a term is a tuple of natural numbers, with $\varepsilon$
denoting the empty tuple. {\em Subterms} and positions are inductively defined
as follows. The term $t$ is a subterm of itself at position $\varepsilon$. If
$s$ is a subterm of $t_i$ at position $p$, then $s$ is a subterm of
$\cst{f}(\tuplen{t})$ at position $i.p$ (with $.$ denoting prepending operation). A {\em context} is a term with zero or
more subterms replaced by a hole $\square$. We write $C[\overline{u}_n]$ for the
term resulting from filling in the holes of a context $C$ with the terms
$\overline{u}_n$, from left to right. We say a term is {\em ground} if it has no
variables. By $u[s]$ we denote that $s$ is a subterm of $u$.

\looseness=-1
{\em Substitutions} $\sigma, \varrho, \ldots$ are total mappings from variables to
terms of the same type. Substitutions map only finitely many variables to terms
other than the variable itself. This is denoted as $\{ x_1 \mapsto t_1, \ldots,
x_n \mapsto t_n \}$ where $x_i$ are variables that are not mapped to themselves.
Applying a substitution $\sigma$ to a term $t$, denoted $\sigma(t)$, results in
replacing all variables of $t$ by the corresponding values of the mapping $\sigma$.
The composition $\varrho\sigma$ of substitutions is defined by
$\left(\varrho\sigma\right)t=\varrho\left(\sigma (t)\right)$. We call a substitution $\sigma$ {\em grounding}
if $\sigma(x) \not= x$ implies that $\sigma(x)$ is ground.

@@

\[
\namedsimp{RN}{s \eq t \qquad u \noteq v \llor D}
              {s \eq t \qquad u[p \leftarrow \sigma(t)] \noteq v  \llor D}
\]
where $\sigma(s) = u|_p$ and $\sigma(s) \succ \sigma(t)$.
